---
title: "Recitation Week 03"
subtitle: "Log-prior and Log-posterior functions: Programming"
author: "Dr. Joseph P. Yurko"
date: "1/27/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, load_packages}
library(dplyr)
library(ggplot2)
```


Let's generate some random numbers.  

```{r, create_random_values}
mu_true <- 254.5
sigma_true <- 2.5

set.seed(5001)
x <- rnorm(n = 50, mean = mu_true, sd = sigma_true)
```

The log-prior.  

```{r, create_logprior}
my_logprior <- function(theta, my_info)
{
  # unknown mean is the first element in theta
  lik_mu <- theta[1]
  # the unkonwn noise is the second element
  lik_sigma <- theta[2]
  
  # calculate the log-prior densities
  dnorm(x = lik_mu,
        mean = my_info$mu_0,
        sd = my_info$tau_0,
        log = TRUE) + 
    dunif(x = lik_sigma,
          min = my_info$sigma_lwr,
          max = my_info$sigma_upr,
          log = TRUE)
}
```

Create a grid of points between the two unknowns.  

```{r, make_parameter_grid}
param_grid <- expand.grid(mu = seq(245, 265, length.out = 201),
                          sigma = seq(0.5, 20.5, length.out = 201),
                          KEEP.OUT.ATTRS = FALSE,
                          stringsAsFactors = FALSE) %>% 
  as.data.frame() %>% tbl_df()
```

Specify the list of required information for the function.  

```{r, create_param_list}
hyper_list <- list(
  mu_0 = 250,
  tau_0 = 2,
  sigma_lwr = 0.5,
  sigma_upr = 20.5
)

hyper_list
```

Define a wrapper function to call the log-density evaluation.  

```{r}
eval_logdensity <- function(mu_val, sigma_val, my_logdens, my_info)
{
  my_logdens(c(mu_val, sigma_val), my_info)
}
```

In `R` there are the suite of `apply` functions. The `tidy` analogs are in the `purrr`.  

```{r}
log_prior_result <- purrr::map2_dbl(param_grid$mu,
                                    param_grid$sigma,
                                    eval_logdensity,
                                    my_logdens = my_logprior,
                                    my_info = hyper_list)
```

purrr::map_dbl(), purrr::map2_dbl(), purrr::pmap_dbl()

Visualize the log-prior surface.  

```{r}
param_grid %>% 
  mutate(log_dens = log_prior_result) %>% 
  mutate(log_dens_2 = log_dens - max(log_dens)) %>% 
  ggplot(mapping = aes(x = mu, y = sigma)) +
  geom_raster(mapping = aes(fill = log_dens_2)) +
  stat_contour(mapping = aes(z = log_dens_2),
               breaks = log(c(0.01/100, 0.01, 0.1, 0.5, 0.9)),
               size = 2.2, 
               color = "black") +
  scale_fill_viridis_c(guide = FALSE,
                       limits = log(c(0.01/100, 1.0))) +
  labs(x = expression(mu), y = expression(sigma)) +
  theme_bw()
```

Now let's create the posterior.  

```{r}
my_logpost <- function(theta, my_info)
{
  # unknown mean is the first element in theta
  lik_mu <- theta[1]
  # the unkonwn noise is the second element
  lik_sigma <- theta[2]
  
  # log-likelihood
  log_lik <- sum(dnorm(x = my_info$xobs,
                       mean = lik_mu,
                       sd = lik_sigma,
                       log = TRUE))
  
  # calculate the log-prior densities
  log_prior <- dnorm(x = lik_mu,
        mean = my_info$mu_0,
        sd = my_info$tau_0,
        log = TRUE) + 
    dunif(x = lik_sigma,
          min = my_info$sigma_lwr,
          max = my_info$sigma_upr,
          log = TRUE)
  
  # add the log-lik to the log-prior
  log_lik + log_prior
}
```

Create a new list which includes the observed values.  

```{r}
info_N05 <- list(
  xobs = x[1:5],
  mu_0 = 250,
  tau_0 = 2,
  sigma_lwr = 0.5,
  sigma_upr = 20.5
)
```

Evaluate the log-posterior over the grid.  

```{r}
log_post_result <- purrr::map2_dbl(param_grid$mu,
                                    param_grid$sigma,
                                    eval_logdensity,
                                    my_logdens = my_logpost,
                                    my_info = info_N05)
```

Visualize the log-posterior.  

```{r}
param_grid %>% 
  mutate(log_dens = log_post_result) %>% 
  mutate(log_dens_2 = log_dens - max(log_dens)) %>% 
  ggplot(mapping = aes(x = mu, y = sigma)) +
  geom_raster(mapping = aes(fill = log_dens_2)) +
  stat_contour(mapping = aes(z = log_dens_2),
               breaks = log(c(0.01/100, 0.01, 0.1, 0.5, 0.9)),
               size = 2.2, 
               color = "black") +
  scale_fill_viridis_c(guide = FALSE,
                       limits = log(c(0.01/100, 1.0))) +
  labs(x = expression(mu), y = expression(sigma)) +
  theme_bw()
```

Optimization to find the mode.  

```{r}
map_result <- optim(c(250, 10),
                    my_logpost,
                    gr = NULL,
                    info_N05,
                    method = "BFGS",
                    hessian = TRUE,
                    control = list(fnscale = -1, maxit = 1001))

map_result
```

