---
title: "Recitation Week 03"
subtitle: "Log-prior and Log-posterior functions: Programming"
author: "Dr. Joseph P. Yurko"
date: "1/27/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load the packages.  

```{r, load_packages}
library(dplyr)
library(ggplot2)
```

Generate the data.  

```{r}
mu_true <- 254.5
sigma_true <- 2.5

set.seed(5001)
x <- rnorm(n = 50, mean = mu_true, sd = sigma_true)
```

Create a function for the log-prior.  

```{r}
my_logprior <- function(theta, my_info)
{
  lik_mu <- theta[1]
  lik_sigma <- theta[2]
  
  dnorm(x = lik_mu,
        mean = my_info$mu_0,
        sd = my_info$tau_0,
        log = TRUE) + 
    dunif(x = lik_sigma,
          min = my_info$sigma_lwr,
          max = my_info$sigma_upr,
          log = TRUE)
}
```

Evaluate the log-prior over a grid.  

```{r}
param_grid <- expand.grid(mu = seq(245, 265, length.out = 201),
                          sigma = seq(0.5, 20.5, length.out = 201),
                          KEEP.OUT.ATTRS = FALSE,
                          stringsAsFactors = FALSE) %>% 
  as.data.frame() %>% tbl_df()
```

Define the list of hyperparameters.  

```{r}
hyper_list <- list(
  mu_0 = 250,
  tau_0 = 2,
  sigma_lwr = 0.5,
  sigma_upr = 20.5
)

hyper_list
```

Create a wrapper function to evaluate the log-prior.  

```{r}
eval_logdensity <- function(mu_val, sigma_val, func_logdens, my_info)
{
  func_logdens(c(mu_val, sigma_val), my_info)
}
```

Calculate the log-prior over the grid of points. Use the `purrr` the basic syntax is `purrr::map_dbl(<input>, <call function>)`. To iterate over 2 things, `purrr::map2_dbl()`. For more than 2 things use `purrr::pmap_dbl()`.  

```{r}
log_prior_result <- purrr::map2_dbl(param_grid$mu,
                                    param_grid$sigma,
                                    eval_logdensity,
                                    func_logdens = my_logprior,
                                    my_info = hyper_list)
```

Visualize the log-prior.  

```{r}
param_grid %>% 
  mutate(log_dens = log_prior_result) %>% 
  mutate(log_dens_2 = log_dens - max(log_dens)) %>% 
  ggplot(mapping = aes(x = mu, y = sigma)) +
  geom_raster(mapping = aes(fill = log_dens_2)) +
  stat_contour(mapping = aes(z = log_dens_2),
               breaks = log(c(0.01/100, 0.01, 0.1, 0.5, 0.9)),
               size = 2.2,
               color = "black") +
  scale_fill_viridis_c(guide = FALSE,
                       limits = log(c(0.01/100, 1.0))) +
  labs(x = expression(mu), y = expression(sigma)) +
  theme_bw()
```

the log-posterior function.  

```{r}
my_logpost <- function(theta, my_info)
{
  lik_mu <- theta[1]
  lik_sigma <- theta[2]
  
  # calculate the log-likelihood
  log_lik <- sum(dnorm(x = my_info$xobs,
                       mean = lik_mu,
                       sd = lik_sigma,
                       log = TRUE))
  
  log_prior <- dnorm(x = lik_mu,
        mean = my_info$mu_0,
        sd = my_info$tau_0,
        log = TRUE) + 
    dunif(x = lik_sigma,
          min = my_info$sigma_lwr,
          max = my_info$sigma_upr,
          log = TRUE)
  
  # sum together
  log_lik + log_prior
}
```

Visualize the log-posterior.  

```{r}
info_N05 <- list(
  xobs = x[1:5],
  mu_0 = 250,
  tau_0 = 2,
  sigma_lwr = 0.5,
  sigma_upr = 20.5
)

log_post_result <- purrr::map2_dbl(param_grid$mu,
                                    param_grid$sigma,
                                    eval_logdensity,
                                    func_logdens = my_logpost,
                                    my_info = info_N05)
```

```{r}
param_grid %>% 
  mutate(log_dens = log_post_result) %>% 
  mutate(log_dens_2 = log_dens - max(log_dens)) %>% 
  ggplot(mapping = aes(x = mu, y = sigma)) +
  geom_raster(mapping = aes(fill = log_dens_2)) +
  stat_contour(mapping = aes(z = log_dens_2),
               breaks = log(c(0.01/100, 0.01, 0.1, 0.5, 0.9)),
               size = 2.2,
               color = "black") +
  scale_fill_viridis_c(guide = FALSE,
                       limits = log(c(0.01/100, 1.0))) +
  labs(x = expression(mu), y = expression(sigma)) +
  theme_bw()
```

