---
title: "Recitation week 02: Programming intro"
author: "Dr. Joseph P. Yurko"
date: "1/13/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Big section

## subsection

### sub-subsection

## Load packages

```{r, load_packages, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
```

## Practice dataset

Work with the `mtcars` dataset.  

New line of text. Another sentence.  

```{r, look_at_mtcars}
mtcars %>% glimpse()
```

Look at the data type of a specific variable.  

```{r, check_datatypes}
class(mtcars$mpg)
```

```{r, display_mtcars}
mtcars
```

Create a new column for the row name.  

```{r, check_class_mtcars}
class(mtcars)

mtcars %>% 
  tbl_df() %>% 
  class()
```

```{r}
mtcars %>% 
  tbl_df() %>% 
  tibble::rowid_to_column() %>% 
  tibble::rownames_to_column("my_rowname") %>% 
  mutate(car_name = rownames(mtcars))
```

```{r}
mtcars
```

Assign our actions to a new variable or object.  

```{r, create_b_dataset}
mtcars_b <- mtcars %>% 
  tbl_df() %>% 
  tibble::rowid_to_column() %>% 
  tibble::rownames_to_column("my_rowname") %>% 
  mutate(car_name = rownames(mtcars))
```

`select()` verb.  

```{r}
mtcars_b %>% 
  dplyr::select(car_name)
```

```{r}
mtcars_b %>% 
  select(car_name) %>% 
  class()
```

```{r}
mtcars_b %>% 
  select(my_rowname, rowid, car_name)
```

```{r}
mtcars_b %>% 
  select(1:3)
```

```{r}
mtcars_b %>% 
  select(c("rowid", "car_name", "my_rowname"))
```

```{r}
mtcars_b %>% 
  select(contains("row"))

mtcars_b %>% 
  select(starts_with("m"))

mtcars_b %>% 
  select(ends_with("g"))
```

## Filter

Use the `filter()` verb to subset rows.  

```{r}
mtcars_b %>% 
  filter(mpg > 20)
```

```{r}
mtcars_b %>% 
  filter(mpg > 20,
         disp > 100)
```

```{r}
mtcars_b %>% 
  filter(mpg > 20 & disp > 100)
```

Use the `arrange()` to rank or sort by a variable.  

```{r}
mtcars_b %>% 
  arrange(mpg)
```

```{r}
mtcars_b %>% 
  arrange(desc(mpg))
```

## Counts

```{r}
mtcars_b %>% 
  distinct(cyl)
```

```{r}
mtcars_b %>% 
  distinct(cyl) %>% 
  arrange(desc(cyl))
```

```{r}
mtcars_b %>% 
  count(cyl)
```

```{r}
mtcars_b %>% 
  group_by(cyl)
```

```{r}
mtcars_b %>% 
  group_by(cyl) %>% 
  summarise(num_rows = n(),
            avg_mpg = mean(mpg))
```

